---
title: "20190627_DiffSplicingBloodSf3b1_MostlyMarrow"
author: "Ben Fair"
date: "6/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```


```{r libraries}
library(data.table)
library(dplyr)
library(ggplot2)
library(reshape2)
library(readxl)
library(ggrepel)
library(tidyverse)
library(gplots)
library(knitr)
library("RColorBrewer")
```



```{r}
#read-count-table

Numers <- '../output/snakemake_out_sf3b1_mostlymarrow/CountTableNumerators.gz'
Denoms <- '../output/snakemake_out_sf3b1_mostlymarrow/CountTableDenominators.gz'

NumeratorCountTable <- fread(paste0("gunzip -c ", Numers), sep=" ", header=TRUE, data.table=FALSE)
DenominatorCountTable <- fread(paste0("gunzip -c ", Denoms), sep=" ", header=TRUE, data.table=FALSE)

PSI_Table <- NumeratorCountTable[-1]/DenominatorCountTable[-1]
row.names(PSI_Table) <- gsub(":clu_.+", "", NumeratorCountTable$chrom)

#Replace NaNs from 0/0 operation
PSI_Table[is.na(PSI_Table)] <- 0

Groups <- read.table("../data/other_snakemake_data/MostlyMarrowContrast.groups", stringsAsFactors = T)
GroupLabels <- plyr::mapvalues(colnames(PSI_Table), from=Groups$V1, to=Groups$V2, warn_missing = FALSE)


#read-leafcutter differntial splicing results
effect.sizes <- read.table(gzfile("../output/snakemake_out_sf3b1_mostlymarrow/leafcutter.ds.effect_sizes.txt.gz"), header=T)

cluster.significance <- fread(paste0("gunzip -c ", "../output/snakemake_out_sf3b1_mostlymarrow/leafcutter.ds.cluster_sig.txt.gz"), sep="\t", header=TRUE, data.table=FALSE)

leafcutter.merged <- effect.sizes %>%
  mutate(cluster=gsub("(.+?:).+?:.+?:(clu.+?)", "\\1\\2", intron, perl=T)) %>%
  mutate(junc_id=gsub("(.+?:.+?:.+?):clu.+", "\\1", intron, perl=T)) %>%
  left_join(cluster.significance, by="cluster")

kable(head(leafcutter.merged))
```


Ok now let's check that the leafcutter differential splicing results make sense. Some diagnositic plots:

```{r}

#scatter plot of two samples
ggplot(data=NULL, aes(x=PSI_Table[,1], y=,PSI_Table[,2])) + geom_point(alpha=0.05)
ColSideColors=HeatmapGroupLabels

#correlation matrix of all samples
cor(PSI_Table, method="spearman") %>%
  heatmap.2(dendrogram="column", trace="none", ColSideColors=GroupLabels)

# how many junctions total for case and control
data.frame(Sum=colSums(NumeratorCountTable[-1])) %>%
  rownames_to_column() %>%
  left_join(Groups, by=c("rowname"="V1")) %>%
  ggplot(aes(x=V2, y=log10(Sum))) +
    geom_boxplot(outlier.shape=NA) + #avoid plotting outliers twice
    geom_jitter(position=position_jitter(width=.2, height=0)) +
    theme_bw()
  


# how many clusters are there
leafcutter.merged %>%
  distinct(cluster, .keep_all = T) %>% dim()

# and how many are significant (padjust<0.05, 0.01)
leafcutter.merged %>%
  distinct(cluster, .keep_all = T) %>% 
  filter(p.adjust<0.1) %>% dim()

#histogram of pvalues
leafcutter.merged %>%
  distinct(cluster, .keep_all = T) %>% 
  ggplot(aes(x=p)) +
    geom_histogram() +
    theme_bw()
```




```{r}
#volcano plot of cluster-pvalues and largest within-cluster-log effect size
leafcutter.merged %>%
  mutate(abs.logef = abs(logef)) %>%
  group_by(cluster) %>%
  slice(which.max(abs.logef)) %>%
  mutate(sig=p.adjust<0.05) %>%
  ggplot(aes(x=logef, y=-log10(p), color=sig)) +
    geom_point(alpha=0.05) +
    scale_x_continuous(limits=c(-10,10)) +
    theme_bw()
```

Same plot but for delta-PSI instead of log-effect size...


```{r}
# volcano plot of cluster-pvalues and largest within-cluster-delta-psi
leafcutter.merged %>%
  mutate(abs.deltapsi = abs(deltapsi)) %>%
  group_by(cluster) %>% 
  slice(which.max(abs.deltapsi)) %>% 
  mutate(sig=p.adjust<0.05) %>%
  ggplot(aes(x=deltapsi, y=-log10(p), color=sig)) +
    geom_point(alpha=0.05) +
    theme_bw()
```

Now let's check that the junctions identified in Darman et al are enriched for small P-values


```{r}
DarmanJunctionsList <- read.table('../data/other_snakemake_data/DarmanDiffJunctions.LiftedToHg38.bed', header=F)

# Make new id, for easier use in later filtering. Note that coordinates may need to add 1
DarmanJunctionsList$junc_id <- paste(DarmanJunctionsList$V1, DarmanJunctionsList$V2 + 1, DarmanJunctionsList$V3 + 1, sep=":")

NotInDarman<-leafcutter.merged %>% 
  group_by(cluster) %>%
  filter(!any(junc_id %in% DarmanJunctionsList$junc_id)) %>%
  slice(which.min(p))
InDarman<-leafcutter.merged %>% 
  group_by(cluster) %>%
  filter(any(junc_id %in% DarmanJunctionsList$junc_id)) %>%
  slice(which.min(p))

## QQ-plot of cluster P-vals, grouped by whether cluster contains in intron identified as differentially spliced in Daraman et al.
ggplot(NotInDarman, aes(color="NotInDarman", y=-log10(sort(p)), x=-log10(1:length(p)/length(p)))) +
  geom_point() +
  geom_point(data=InDarman, aes(color="InDarman")) +
  xlab("-log10(Theoretical-Pvalues)") +
  ylab("-log10(Observed-Pvalues)") +
  geom_abline() +
  theme_bw() +
  theme(legend.position="bottom") +
  theme(legend.title=element_blank())
```


PCA and heatmap of differentially spliced introns

```{r}
#Get top intron within each cluster (by delta-psi) then subset the top 1000 clusters by Pval. This is the top1000 introns that will be used to plot heatmap of PSI
Top1000Introns <- leafcutter.merged %>%
  mutate(abs.deltapsi = abs(logef)) %>%
  group_by(cluster) %>%
  slice(which.max(abs.deltapsi)) %>%
  ungroup() %>%
  top_n(-1000, p) %>% pull(intron)



#First make a heatmap of Darman et al junctions
Heatmap <- PSI_Table[DarmanJunctionsList$junc_id,] %>%
  t() %>% scale(scale=F) %>% t() %>%
  as.data.frame() %>% drop_na() %>% as.matrix()

heatmap.2(Heatmap, col=colorRampPalette(c("blue", "white", "red"))(n = 299), dendrogram="column", trace="none", labRow="", key.title="PSI-Zscore", ColSideColors=GroupLabels)

#Heatmap of top leafcutter identified introns
Heatmap <- PSI_Table[Top1000Introns,] %>%
  t() %>% scale(scale=F) %>% t() %>%
  as.data.frame() %>% drop_na() %>% as.matrix()
heatmap.2(Heatmap, col=colorRampPalette(c("blue", "white", "red"))(n = 299), dendrogram="column", trace="none", labRow="", key.title="PSI-Zscore", ColSideColors=GroupLabels)
```

What's up with the near-identical samples: check correlation matrix


```{r}
AnnotatedIntrons <- read.table(gzfile("../data/other_snakemake_data/GencodeHg38_all_introns.bed.gz"), sep='\t', col.names = c("chrom", "start", "stop", "gene", "sysID", "strand", "transcriptID", "intronnum", "type", "other")) %>%
  mutate(thress = case_when(
    strand == "+" ~ paste(chrom, stop, strand),
    strand == "-" ~ paste(chrom, start+1, strand))) %>%
  mutate(fivess = case_when(
    strand == "+" ~ paste(chrom, start+1, strand),
    strand == "-" ~ paste(chrom, stop, strand)))

PSI_Table.Extra <- PSI_Table %>% as.data.frame() %>%
  rownames_to_column() %>%
  mutate(junc_name = DenominatorCountTable$chrom) %>%
  separate(junc_name, c("chrom", "start", "stop", "cluster"), sep=":") %>%
  mutate(strand=str_sub(cluster, -1)) %>%
  mutate(thress = case_when(
    strand == "+" ~ paste(chrom, stop, strand),
    strand == "-" ~ paste(chrom, start, strand))) %>%
  mutate(fivess = case_when(
    strand == "+" ~ paste(chrom, start, strand),
    strand == "-" ~ paste(chrom, stop, strand)))
  
Crytpic3ss <- PSI_Table.Extra %>%
  filter((fivess %in% AnnotatedIntrons$fivess) & (!thress %in% AnnotatedIntrons$thress)) %>%
  pull(rowname)

Cryptic3ssQuant <- NumeratorCountTable %>%
  mutate(junc_id = gsub(":clu_.+", "", chrom)) %>%
  filter(junc_id %in% Crytpic3ss) %>%
  select(-chrom, -junc_id) %>% colSums()

TotalSum <- NumeratorCountTable %>% select(-chrom) %>% colSums()
FractionCryptics3ss.df <- data.frame(TotalSum,Cryptic3ssQuant) %>%
  rownames_to_column() %>%
  mutate(sf3b1status = plyr::mapvalues(rowname, from=Groups$V1, to=Groups$V2, warn_missing = FALSE)) %>%
  mutate(FractionCryptic3ss = Cryptic3ssQuant/TotalSum)
ggplot(FractionCryptics3ss.df, aes(y=FractionCryptic3ss, x=sf3b1status)) +
  geom_boxplot(outlier.shape=NA) + #avoid plotting outliers twice
  geom_jitter(position=position_jitter(width=.2, height=0)) +
  theme_bw()

wilcox.test(FractionCryptic3ss ~ sf3b1status, data=FractionCryptics3ss.df)

####
Top1000Quant <- NumeratorCountTable %>%
  filter(chrom %in% Top1000Introns) %>%
  select(-chrom) %>% colSums()

TotalSum <- NumeratorCountTable %>% select(-chrom) %>% colSums()
FractionTop1000.df <- data.frame(TotalSum,Top1000Quant) %>%
  rownames_to_column() %>%
  mutate(sf3b1status = plyr::mapvalues(rowname, from=Groups$V1, to=Groups$V2, warn_missing = FALSE)) %>%
  mutate(FractionCryptic3ss = Top1000Quant/TotalSum)
ggplot(FractionTop1000.df, aes(y=FractionCryptic3ss, x=sf3b1status)) +
  geom_boxplot(outlier.shape=NA) + #avoid plotting outliers twice
  geom_jitter(position=position_jitter(width=.2, height=0)) +
  theme_bw()

wilcox.test(FractionCryptic3ss ~ sf3b1status, data=FractionCryptics3ss.df)
  
```

